Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-09-15 iso-int/rc-iso-int.tex
many small edits

\subsection{Universality.} 
As an example, consider the implementation of the Toffoli gate below,
which establishes that {{langRevT}} is complete for combinational
circuits. This encoding uses the type {{1+1}} as {{bool}} and values
{{left ()}} and {{right()}} as {{true}} and {{false}}.  Boolean
negation not is simply {{swap+}}. The Toffoli gate takes three boolean
inputs: if the first two inputs are {{true}} then the third is
negated.

Even though {{langRevT}} lacks conditional expressions, they are
expressible using the distributivity laws as we demonstrate.  Given
any combinator {{c : b <-> b}} we can construct a combinator called
{{if_c : bool*b <-> bool*b}} in terms of {{c}}, where {{if_c}}
behaves like a one-armed if-expression. If the supplied boolean is
true then the combinator {{c}} is used to transform the value of type
{{b}}. If the boolean is {{false}}, then the value of type {{b}}
remains unchanged. We can write down the combinator for {{if_c}} in
terms of {{c}} as {{distrib (;) ((id (*) c) (+) id) (;) factor}}.

Given {{if_c}}, constructing the Toffoli gate is straightforward. If
we choose {{c}} to be {{not}} we get {{if_{not} }} which is
{{cnot:bool*bool<->bool*bool}} and {{if_{cnot} }} is the required
{{toffoli : bool * (bool * bool) <-> bool * (bool * bool)}}, drawn
below.
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/toffoli.pdf}
}
\end{center}
\noindent Previous work~\cite[Sec.~5]{James:2012:IE:2103656.2103667}
establishes that {{langRevT}} is Turing complete by showing the compilation
of a Turing complete language --- a first-order language with numbers and
iteration --- to {{langRevT}}.

@@

In {{langRevT}} we encode numbers, {{nat = 0 ~|~ n + 1}}, using the
recursive type {{@@x.1+x}}. The {{fold}}/{{unfold}} isomorphisms for
{{nat}} are {{unfold : nat <-> 1+nat : fold}}.
\begin{multicols}{3}
\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/unfold_nat.pdf}
}
\end{center}

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/fold_nat.pdf}
}
\end{center}

\begin{center}
\scalebox{0.8}{
  \includegraphics{diagrams/add1.pdf}
}
\end{center}
\end{multicols}
The triangle on the left denotes the {{unfold}} isomorphism, which
works as follows: If the number {{n}} is zero, the output is the top
branch which has type~{{1}}. If the number is non-zero, the output is
on the bottom branch and has value {{n-1}}. The triangle in the middle
is the dual {{fold}} isomorphism.

As explained in previous work
\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}, numeric addition and
subtraction can be expressed in {{langRevT}} as partial
isomorphisms. The combinator on the right above is
{{add1:nat<->nat}}. It acts like the successor function returning
{{n+1}} for all inputs {{n}}. Its adjoint {{sub1:nat<->nat}}, which we
get by flipping the diagram right to left, diverges on input {{0}} and
returns {{n-1}} for all other inputs {{n}}.

@@

Deriving {{langRevT}} combinators from abstract machines can sometimes
be used as an efficient indirect way of programming in {{langRevT}}.
It is well known that every number {{n}} can be represented as
{{2a+0}} or {{2a+1}} depending on whether it is odd or even.  The
later can be viewed as the algebraic data type
{{par=odd~|~even~|~A~par}} such that 
{{0 = even, 1 = odd, 2 = A even, 3 = A odd, 4 = A (A even)}} etc where
the nesting of {{A}} constructor indicates the value of {{a}}.

Say we wish to build a {{langRevT}} combinator to map {{nat}} to its
parity encoded version {{par}}, expressible as the {{langRevT}} type
{{@@x.(1+1)+x}}. Since {{par}} is not a fixed unfolding of {{nat}}
({{@@x.1+x}}), it is not immediately apparent how such a combinator
can be constructed.

Instead of directly programming in {{langRevT}}, one can derive the
required {{nat<->par}} combinator by first constructing an abstract
machine that maps {{nat}} to {{par}}. We first express {{nat}} and
{{par}} algebraically and then develop a logically reversible abstract
machine expressing the translation:

@@

The construction of the {{langRevT}} combinator is
straightforward and proceeds as before. The partial combinator
representing the wiring of machine reductions is shown below.

>>>>

\subsection{Universality.} 
As an example, consider the implementation of the Toffoli gate below,
which establishes that {{langRevT}} is complete for combinational
circuits. The Toffoli gate takes three boolean inputs: if the first
two inputs are {{true}} then the third is negated. This encoding uses
the type {{1+1}} as {{bool}} and values {{left ()}} and {{right()}} as
{{true}} and {{false}}.  Boolean negation, {{not : bool<->bool}}, is
simply {{swap+}}.

Even though {{langRevT}} lacks conditional expressions, they are
expressible using the distributivity laws.  Given any combinator
{{c:b<->b}}, we can construct a combinator called
{{if_c:bool*b<->bool*b}} in terms of {{c}}, which behaves
like a one-armed if-expression. If the supplied boolean is {{true}} then
the combinator {{c}} is used to transform the value of type {{b}}. If
the boolean is {{false}} then the value of type {{b}} remains
unchanged. We can write down the combinator for {{if_c}} in terms of
{{c}} as {{distrib (;) ((id (*) c) (+) id) (;) factor}}.

Given {{if_c}}, constructing the Toffoli gate is straightforward. If
we choose {{c}} to be {{not}} we get {{if_{not} }} which is controlled-not,
{{cnot:bool*bool<->bool*bool}}. Further, {{if_{cnot} }} is the required
Toffoli gate, 
{{toffoli : bool * (bool * bool) <-> bool * (bool * bool)}}, and is drawn
below.
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/toffoli.pdf}
}
\end{center}
\noindent Previous work~\cite[Sec.~5]{James:2012:IE:2103656.2103667}
establishes that {{langRevT}} is Turing complete by showing the compilation
of a Turing complete language --- a first-order language with numbers and
iteration --- to {{langRevT}}.

@@

We encode numbers, {{nat = 0 ~|~ n + 1}}, using the
recursive {{langRevT}} type {{@@x.1+x}}. The {{fold}}/{{unfold}} isomorphisms for
{{nat}} are {{unfold : nat <-> 1+nat : fold}}.
\begin{multicols}{3}
\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/unfold_nat.pdf}
}
\end{center}

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/fold_nat.pdf}
}
\end{center}

\begin{center}
\scalebox{0.9}{
  \includegraphics{diagrams/add1.pdf}
}
\end{center}
\end{multicols}
The combinator on the left denotes the {{unfold}} isomorphism, which
works as follows: If the number {{n}} is zero, the output is the top
branch which has type~{{1}}. If the number is non-zero, the output is
on the bottom branch and has value {{n-1}}. The combinator in the middle
is its dual {{fold}} isomorphism.

As explained in previous work
\cite[Sec.~4.2]{James:2012:IE:2103656.2103667}, numeric addition and
subtraction can be expressed in {{langRevT}} as partial
isomorphisms: The combinator on the right, {{add1:nat<->nat}}, acts
like the successor function returning {{n+1}} for all inputs
{{n}}. Its adjoint, {{sub1}}, obtained by flipping the
diagram right to left, is a partial map that diverges on input {{0}}
and returns {{n-1}} for all other inputs {{n}}.

@@

Deriving {{langRevT}} combinators from abstract machines can sometimes
be used as an efficient indirect way to program in {{langRevT}}.  It
is well known that every number {{n}} can be represented as {{2a+0}}
or {{2a+1}}, depending on whether it is odd or even.  The later can be
represented by the algebraic data type {{par=odd~|~even~|~A~par}}
where the nesting of {{A}} constructor indicates the value of
{{a}}. For example 
{{0 = even, 1 = odd, 2 = A even, 3 = A odd, 4 = A (A even)}} etc.

Say we wish to build a {{langRevT}} combinator to map {{nat}} to its
parity encoded version {{par}}. While the type {{par}} is expressible
as {{@@x.(1+1)+x}} in {{langRevT}}, it is not a fixed unfolding of
{{nat = @@x.1+x}} and hence it is not immediately apparent how such a
combinator can be constructed.

Instead of directly programming in {{langRevT}} however, one can
derive the required {{nat<->par}} combinator by first constructing an
abstract machine that maps {{nat}} to {{par}} and then translating it
to {{langRevT}}. We first express {{nat}} and {{par}} algebraically
and then develop a logically reversible abstract machine:

@@

The derivation of the {{langRevT}} combinator is
straightforward and proceeds as before. The partial combinator
representing the wiring of machine reductions is shown below.
